Realizing temporal graphs from fastest travel times

### 2023-02-09 ICALP23-RealizingTemporalGraphs.tex
Final versions_001

\usepackage{thmtools} 
\usepackage{thm-restate}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% PROBLEMS NAMES
%% One label per edge, period Delta, exact realization
\newcommand{\deltaExactLong}{\textsc{Simple periodic Temporal Graph Realization}}
\newcommand{\deltaExact}{\textsc{Simple TGR}}

@@

\bibliographystyle{plainurl}% the mandatory bibstyle

@@

\begin{abstract}
In this paper we study \emph{temporal design} problems of undirected temporally connected graphs. 
The basic setting of these optimization problems is as follows: given an undirected graph~$G$, 
what is the smallest number $|\lambda|$ of time-labels that we need to add to the edges of $G$ such that the resulting temporal graph $(G,\lambda)$ is temporally connected?

@@

Graph realization problems are fundamental graph-algorithmic questions where the objective is to (re)construct a graph, 
given some constraints on some of its graph attributes. Among the most well studied such parameters for graph reconstruction 
are the vertex degrees~\cite{cai2000solution,rfgh611961hu,hakimi1962realizability,barNoy2019graphRealizations,barNoy2021selected,erdos1960graphs}, distances~\cite{barNoy2022GraphRealization,barNoy2021composed,hakimi1965distance}, eccentricities~\cite{barNoy2020efficiently,hell2009linear,behzad1976eccentric,lesniak1975eccentric}, connectivity~\cite{fulkerson1960zero,frank1992augmenting,chen1966realization,frank1994connectivity,frank1970connectivity,gomory1961multi}, etc. 
The motivation for network realization problems stems from both ``verification'' applications where we attempt to reconstruct an existing network of unknown structure based on the outcomes of experimental measurements, and engineering applications related to network design~\cite{augustine2022distributed,grotschel1995design}.

Due to space constraints, proofs of results marked with $\star$ are (partially) deferred to Appendix.

SKELETON:
\begin{itemize}
    \item graph realization problems in general, well studied and motivated for different types of graph attributes (among the most natural ones are distances and degree sequences), mention the two perspectives: verification and network design. 
    \item Some observations and references about distance realization known results on static graphs.
    \item In modern network applications, network structure changes over time
    \item In this paper we focus on distance attributes (one of the most fundamental graph attributes)
    \item What is distances in temporal graphs? Fastest, etc, periodic
    \item Our contributions
    \item Related work on temporal network design etc
\end{itemize}

Sections for ICALP:
\begin{enumerate}
    \item Intro 2 PAGES (GEORGE + PAUL)
    \item Prelims 1 PAGE
    \item Hardness (HENDRIK SHORTENS)
    \begin{enumerate}
        \item NP 1/2 PAGE (APPENDIX)
        \item W1 (3 PAGES)
    \end{enumerate}
    \item Algos (NINA SHORTENS)
    \begin{enumerate}
        \item Tree ALGO 2 PAGE
        \item FPT ALGO (6 PAGES)
        \item Cylce ALGO (APPENDIX)
    \end{enumerate}
    \item Conclusion (?) 1/2 PAGE
\end{enumerate}

\section{Motivation}

	When considering the problem of finding shortest paths among vertices, one relies heavily on the optimal substructure of the problem,
	which states that if a shortest path $P$ from vertex $u$ to $v$ traverses the vertex $w$,
	then the shortest path from $u$ to $w$ is a subpath of $P$.
	After studying the problem of fastest temporal paths (\ie temporal paths of smallest duration),
	one can see that this property does not carry over
	(for an example see~\cref{fig:ftpExample}).
	Which makes all the standard intuition and work with these kind of paths, much harder.
    This also explains why our FPT algorithm, w.r.t the feedback edge number, is so complex. Usually, the feedback edge number parameter (denote it $k$) provides an easy algorithm, 
    where one first determines $O(k)$ special vertices,
    guesses desired structures among them (\ie iterates through all possibilities) and then simply extends the solution to the rest of the graph.
    An example with this approach, on temporal graphs, can be observed in \cite{casteigts2021finding} where the authors provide an FPT algorithm that determines all temporal paths, under certain constraints.
    In our case the similar approach cannot work, as we cannot at any point calculate some partial labeling $\lambda$ of a delta periodic graph $(G,\lambda)$,
    that could at the end be extended to the whole $G$.

	\begin{figure}[htb]
    \centering
	\begin{tikzpicture}[xscale=2]
	%%%S_uv
	\node[vert,label=below:$u$] (v1) at (1,0) {};
	\node[vert] (v2) at (2,0) {};
	\node[vert,label=below:$w$] (v3) at (3,0) {};
	\node[vert] (v4) at (4,0) {};
	\node[vert,label=below:$v$] (v5) at (5,0) {};
	\draw (v1) -- node[timelabel] {$1$} (v2) -- node[timelabel] {$3$}  (v3) -- node[timelabel] {$5$} (v4) -- node[timelabel] {$7$} (v5);
	\draw (v1) to [out=80,in=100,distance=2cm] node[timelabel] {$9$} (v3);

\section{Preliminaries}
An interval on $\mathbb N$ from $a$ to $b$ is denoted by $[a,b] \coloneqq \{ i\in \mathbb N_0 \mid a \leq i \leq b\}$
and $[a] \coloneqq [1,a]$.
An undirected graph~$G=(V,E)$ consists of a set~$V$ of vertices 
and a set~$E \subseteq V \times V$ of edges.
For a graph~$G$, we also denote by~$V(G)$ and~$E(G)$ the vertex and edge set of~$G$, respectively.
We denote an edge $e \in E$ between vertices $u,v \in V$ as a set $e=\{u,v\}$.
For the simplicity and the sake of an easier representation we sometimes denote an edge $e$ also as a sequence $uv$.
A path~$P \subset G$ is a graph with a vertex set~$V(P)=\{v_1,\dots,v_k\}$ of distinct vertices and edge set~$E(P)=\{\{v_i,v_{i+1}\}\mid 1\leq i<k\}$
(we often represent path~$P$ by the tuple~$(v_1,v_2,\dots,v_k)$).

@@

Let $G=(V,E)$ and $\Delta\in \mathbb{N}$, and let $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ be an edge-labeling function that assigns to every edge of $G$ one of the labels from $\{1,\ldots,\Delta\}$. 
Then we denote by $(G,\lambda,\Delta)$ the \emph{$\Delta$-periodic temporal graph} $(G,\lambda_{\Delta})$, where for every edge $e\in E$ we have $\lambda_{\Delta}(e)=\{i\Delta + x : i\geq 0, x\in \lambda(e)\}$. 
In this case we call $\lambda$ a \emph{$\Delta$-periodic labeling} of $G$.
When it is clear from the context, we drop $\Delta$ form the notation and say that $(G,\lambda)$ is a ($\Delta$-periodic) temporal graph.

\begin{definition}[temporal path] \label{def:temporalPath+Duration}
Let $(G,\lambda)$ be a temporal graph. A \emph{temporal path} $(P,\lambda)$ from $u=u_0$ to $v=u_k$ in $(G,\lambda)$ is a sequence $(u_0 u_1,t_1),(u_1 u_2,t_2),\ldots,(u_{k-1} u_k,t_k)$, 
where $P=(u_0,u_1,\ldots,u_k)$ is a path in the underlying static graph $G$ (we usually say that $P$ is the underlying path of $(P,\lambda)$), 
$t_i\in \lambda_\Delta(u_{i-1}u_i)$ for every $i=1,\ldots,k$, and $t_1<t_2<\ldots<t_k$. 
The \emph{duration} of this temporal path $(P,\lambda)$ from $u$ to $v$ is $d(P,\lambda)=t_k - t_1 + 1$.
A \emph{fastest} temporal path $(P,\lambda)$ from $u$ to $v$ in $(G,\lambda)$ is a temporal path from $u$ to $v$ with the smallest possible duration.
The duration of the \emph{fastest} temporal path from $u$ to $v$ is denoted by $d(u,v)$.
\end{definition}

Let $P$ be a path from $u$ to $v$ in $G$. 
We denote with $d(u,v,P)$ the duration of the fastest temporal path from $u$ to $v$ along path $P$. Note that $d(u,v,P) \geq d(u,v)$.

\problemdef{\deltaExactLong\ (\deltaExact)}
{An $n \times n$ matrix $D$, a positive integer $\Delta$.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ 
and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is equal to the entry $D_{i,j}$.}

Let $v_1,v_2,\ldots,v_n$ be the $n$ vertices of the graph $G$ that we are looking for in the above problems. 
For simplicity of the presentation (and with a slight abuse of notation), we refer during the paper to the entry $D_{i,j}$ of the matrix $D$ as 
$D_{a,b}$, where $a=v_i$ and $b=v_j$. That is, we put as indices of the matrix $D$ the corresponding vertices of $G$ instead of their numbering.
Many times we also refer to a path $P=(u=v_1, v_2, \dots, v_p=v)$ from $u$ to $v$ in $G$,
as a temporal path in $(G,\lambda)$,
where we actually mean that $(P,\lambda)$ is a temporal path with the underlying path $P$.

In their work Wu et~al.~\cite{Wu2016Efficient} provide an algorithm that computes the duration of the fastest temporal path from a source vertex $v$, in a temporal graph $(G,\lambda)$, to all other vertices,
that runs in polynomial-time.
Using their algorithm, given a temporal graph $(G,\lambda)$, one can determine the whole matrix $D$ in polynomial-time,
where $D$ consists of durations of fastest temporal paths among all pairs of vertices in $(G,\lambda)$.

@@

\section{\deltaExact}

In this section 
we show that in the case when the underlying graph $G$ of an instance $(D, \Delta)$ of \deltaExact\ is a tree or a cycle,
we can determine desired $\Delta$-periodic labeling $\lambda$ of $G$ in polynomial time.
We first start with defining certain notions, that will be of use when solving the problem.

    \begin{definition}[Travel delays]
    Let $(G, \lambda, \Delta)$ be a temporal graph satisfying conditions of \deltaExact.
    Let $e_1=uv$ and $e_2=vz$ be two incident edges in $G$ with $e_1 \cap e_2 = v$.
    We define the \emph{travel delay} from $u$ to $z$ at vertex $v$, denoted with $\tau_v^{uz}$,
    as the difference of the labels of $e_2$ and $e_1$, where we subtract the value of the label of $e_1$ from the label of $e_2$,  modulo~$\Delta$.
    More specifically:
    \begin{equation}\label{eq:def-VertexWaitingTime}
       \tau_v^{uz} \equiv \lambda (e_2) - \lambda(e_1) \pmod \Delta.
    \end{equation}
    Similarly, $\tau_v^{zu} \equiv \lambda (e_1) - \lambda(e_2) \pmod \Delta$.
    \end{definition}
    Intuitively, the value of $\tau_v^{uz}$ represents how long a temporal path waits at vertex $v$ when first taking edge $e_1=uv$ and then edge $e_2 = vz$.

    From the above definition and the definition of the duration of the temporal path $P$ we get the following two observations.
    \begin{observation}\label{obs:durationPwithWaitingTimes}
        Let $P = (v_0, v_1, \dots, v_p)$ be the underlying path of the temporal path
        $(P, \lambda)$ from $v_0$ to $v_p$.
        Then $d(P,\lambda) = \sum_{i = 1}^{p-1} \tau_{v_i}^{v_{i-1}v_i} + 1 $.
    \end{observation}
    \begin{proof}
        For the simplicity of the proof denote $t_i = \lambda(v_{i-1}v_i)$, and suppose that $t_i \leq t_{i+1}$, for all $i \in \{1,2,3,\dots,p\}$.
        Then
        \begin{align*}
        \sum_{i = 1}^{p-1} \tau_{v_i}^{v_{i-1}v_i} + 1  
        &= \sum _{i = 1}^{p-1} (t_{i+1} - t_i) + 1 \\
        & = (t_2 - t_1) + (t_3  t_2) + \cdots + (t_p - t_{p-1}) + 1  \\
        & = t_{p-1} - t_1 + 1\\
        & = d(P)
        \end{align*}
        Now in the case when $t_i > t_{i+1}$ we get that $\tau_{v_i}^{v_{i-1}v_{i+1}} = \Delta + t_{i+1} - t_i$.
        At the end this still results in the correct duration as the last time we traverse the path $P$ is not exactly $t_p$ but $k \lambda + t_p$, for some $k$.
    \end{proof}
    We also get the following.
    \begin{observation}\label{obs:travel-delays-both-directions}
    Let $(G, \lambda, \Delta)$ be a temporal graph satisfying conditions of problem \deltaExact.
    For any two incident edges $e_1 = uv$ and $e_2 = vz$ on vertices $u,v,z \in V$, with $e_1 \cap e_2 = v$, we have $\tau_v^{zu} = \Delta - \tau_v^{uz} \pmod \Delta$.
    \end{observation}

\begin{lemma} \label{lemma:unique-fastest-path-allDelays}
    Let $(G, \lambda, \Delta)$ be a temporal graph satisfying conditions of problem \deltaExact,
    and let $(P,\lambda)=(P_{1,k},\lambda)$ be a fastest temporal path from $u=v_1$ to $v=v_k$ on $k$ vertices $v_1,v_2,\dots,v_k$.
    Let us denote with $(P_{1,i}, \lambda)$ the sub-path of temporal path $(P_{1,k},\lambda)$, that starts at $v_1$ and finishes at $v_i$.
    Suppose that $(P_{1,i}, \lambda)$ is also the fastest temporal path from $u=v_1$ to any other vertex $v_i$ in $P$.
    Then we can determine travel delays on vertices of $P$ using the following equation
    \begin{equation}\label{eq:trav-delays-paths}
        \tau_{v_i}^{v_{i-1},v_{i+1}} = D_{1,i+1} - D_{1,i},
    \end{equation}
    for all $i \in \{2,3, \dots, k-1\}$.
\end{lemma}

>>>>

\usepackage{thmtools} 
\usepackage{thm-restate}
\usepackage{cite}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% PROBLEMS NAMES
%% One label per edge, period Delta, exact realization
\newcommand{\deltaExactLong}{\textsc{Periodic Temporal Graph Realization}}
\newcommand{\deltaExact}{\textsc{Periodic TGR}}

@@

\bibliographystyle{plainurl}% the mandatory bibstyle
\usepackage{cite} %order citations by increasing number

@@

\begin{abstract}
In this paper we initiate the study of the \emph{temporal graph realization} problem with respect to the fastest path durations among its vertices, 
while we focus on periodic temporal graphs. 
Given an $n \times n$ matrix $D$ and a $\Delta \in \mathbb{N}$, the goal is to construct a $\Delta$-periodic temporal graph with $n$ vertices 
such that the duration of a \emph{fastest path} from $v_i$ to $v_j$ is equal to $D_{i,j}$, or to decide that such a temporal graph does not exist. 
The variations of the problem on static graphs has been well studied and understood since the 1960's, and this area of research remains active until nowadays.

As it turns out, the periodic temporal graph realization problem has a very different computational complexity behavior than its static (i.e.~non-temporal) counterpart. 
First we show that the problem is NP-hard in general, but polynomial-time solvable if the so-called underlying graph is a tree or a cycle.
Building upon those results, we investigate its parameterized computational complexity with respect to structural parameters of the underlying static graph which measure the ``tree-likeness''. For those parameters, we essentially give a tight classification between parameters that allow for tractability (in the FPT sense) and parameters that presumably do not.
We show that our problem is W[1]-hard when parameterized by the \emph{feedback vertex number} of the underlying graph, while we show that it is in FPT when parameterized by the \emph{feedback edge number} of the underlying graph.

@@

The \emph{graph realization} problem with respect to a graph property $\mathcal{P}$ is to find a graph that satisfies property $\mathcal{P}$, or to decide that no such graph exists. 
The motivation for graph realization problems stems both from ``verification'' applications, 
where we attempt to reconstruct an existing network of unknown structure based on the outcomes of experimental measurements, 
and from network design applications in engineering, where the goal is to design network topologies having a desired property~\cite{augustine2022distributed,grotschel1995design}.
Analyzing the computational complexity of the graph realization problems for various natural and fundamental graph properties $\mathcal{P}$ requires a deep understanding of these properties.
Among the most well studied such parameters for graph realization 
are constraints on the distances between vertices~\cite{barNoy2022GraphRealization,barNoy2021composed,hakimi1965distance,chung2001distance,bixby1988almost,culberson1989fast}, 
on the vertex degrees~\cite{GolovachM17,gomory1961multi,hakimi1962realizability,Bar-NoyCPR20,erdos1960graphs}, 
%%%%barNoy2021selected,cai2000solution
on the eccentricities~\cite{barNoy2020efficiently,hell2009linear,behzad1976eccentric,lesniak1975eccentric}, and on connectivity~\cite{fulkerson1960zero,frank1992augmenting,chen1966realization,frank1994connectivity,frank1970connectivity,gomory1961multi}, among others.

In the simplest version of a graph realization problem with respect to vertex distances, 
we are given a symmetric $n \times n$ matrix $D$ and we are looking for an $n$-vertex undirected and unweighted graph $G$ such that $D_{i,j}$ equals the distance between vertices $v_i$ and $v_j$ in~$G$. This problem can be trivially solved in polynomial time in two steps~\cite{hakimi1965distance}: First, we build the graph $G=(V,E)$ such that $v_i v_j \in E$ if and only if $D_{i,j}=1$. Second, from this graph $G$ we compute the matrix $D_G$ which captures the shortest distances for all pairs of vertices. If $D_G = D$ then $G$ is the desired graph, otherwise there is no graph having $D$ as its distance matrix. 
Non-trivial variations this problem have been extensively studied, such as for weighted graphs~\cite{hakimi1965distance,Patrinos-Hakimi-72}, as well as for cases where the realizing graph has to belong to a specific graph family~\cite{hakimi1965distance, barNoy2021composed}. Other variations of the problem include the cases where every entry of the input matrix $D$ may contain a range of consecutive permissible values~\cite{barNoy2021composed,Rubei16,Tamura93}, or even an arbitrary set of acceptable values~\cite{barNoy2022GraphRealization} for the distance between the corresponding two vertices.

In this paper we make the first attempt to understand the complexity of the graph realization problem with respect to vertex distances in the context of \emph{temporal graphs}, i.e.~of graphs whose \emph{topology changes over time}.

@@

Here, whenever $t \in \lambda(e)$, we say that the edge $e$ is \emph{active} or \emph{available} at time $t$. In the context of temporal graphs, where the notion of vertex adjacency is time-dependent, the notions of path and distance also need to be redefined. The most natural temporal analogue of a path is that of a \emph{temporal} (or \emph{time-dependent}) path, which is motivated by the fact that, due to
causality, entities and information in temporal graphs can ``flow'' only along sequences of
edges whose time-labels are strictly increasing.

\begin{definition}[fastest temporal path] \label{def:temporalPath+Duration}
Let $(G,\lambda)$ be a temporal graph. A \emph{temporal path} 
in $(G,\lambda)$ is a sequence $(e_1,t_1),(e_2,t_2),\ldots,(e_k,t_k)$, 
where $P=(e_1,\ldots,e_k)$ is a path in the underlying static graph $G$, 
%(we usually say that $P$ is the underlying path of $(P,\lambda)$), 
$t_i\in \lambda(e_i)$ for every $i=1,\ldots,k$, and $t_1<t_2<\ldots<t_k$. 
The \emph{duration} of this temporal path %$(P,\lambda)$ from $u$ to $v$ 
is $t_k - t_1 + 1$.
A \emph{fastest} temporal path from a vertex $u$ to a vertex $v$ in $(G,\lambda)$ is a temporal path from $u$ to $v$ with the smallest duration.
The duration of the \emph{fastest} temporal path from $u$ to $v$ is denoted by $d(u,v)$.
\end{definition}

%Other temporal analogues of a shortest path are temporal paths with the smallest total number of edges (\emph{shortest} temporal path), or the smallest earliest arrival time (\emph{foremost} temporal path), respectively.

In an arbitrary temporal graph $(G,\lambda)$ there can be pairs of vertices $u,v$ where there is no temporal path at all from $u$ to $v$, even if the underlying graph $G$ is undirected and connected. 
In order to focus on the smallest duration of temporal graphs, rather than on temporal reachability between vertices, we consider in this paper only \emph{periodic} temporal graphs, i.e.~on temporal graphs in which the availability times of every edge are periodic (all edges having the same period $\Delta$)~\cite{casteigts2012time}. 
Apart from the theoretical interest of periodic temporal graphs, which have not been well studied and understood in the literature, they capture naturally many application areas too, such as transportation or satellite networks, where entities move according to repeating schedules or time tables. For example, a fastest temporal path from vertex $u$ to vertex $v$ in the temporal graph defined by a transportation (resp.~satellite) network represents the shortest travel time (resp.~the shortest communication delay) from $u$ to $v$. 
As it turns out, the periodic temporal graph realization problem with respect to a given $n \times n$ matrix $D$ of the fastest duration times has a very different computational complexity behavior than the classic graph realization problem with respect to shortest path distances in static graphs.

Formally, let $G=(V,E)$ and $\Delta\in \mathbb{N}$, and let $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ be an edge-labeling function that assigns to every edge of $G$ exactly one of the labels from $\{1,\ldots,\Delta\}$. 
Then we denote by $(G,\lambda,\Delta)$ the \emph{$\Delta$-periodic temporal graph} $(G,L)$, where for every edge $e\in E$ we have $L(e)=\{i\Delta + x : i\geq 0, x\in \lambda(e)\}$. 
In this case we call $\lambda$ a \emph{$\Delta$-periodic labeling} of $G$. 
When it is clear from the context, we drop $\Delta$ form the notation and say that $(G,\lambda)$ is a ($\Delta$-periodic) temporal graph.
Given a duration matrix $D$, it is easy to observe that similar to the static case, if $D_{i,j}=1$, then $v_i$ and $v_j$ must be connected by an edge. We call the graph defined by those edges as the \emph{underlying graph} of $D$.

\subparagraph{Our contribution.}
%In this work, 
We initiate the study of naturally motivated graph realization problems in the temporal setting. To this end, we introduce and investigate the following problem, capturing the setting described above:

\problemdef{\deltaExactLong\ (\deltaExact)}
{An $n \times n$ matrix $D$, a positive integer $\Delta$.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ 
and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is $D_{i,j}$?}

We start with showing NP-hardness of the problem (\cref{thm:NPhardness}), even if $\Delta$ is a small constant. To establish a baseline for tractability, we show that \deltaExact\ is polynomial-time solvable if the underlying graph is a tree (\cref{thm:deltaExact-PolyTimeTrees}).

Building upon these initial results, we explore the possibilities to generalize our polynomial-time algorithm using the \emph{distance-from-triviality} parameterization paradigm~\cite{FJR13,GHN04}. That is, we investigate the parameterized computational complexity of \deltaExact\ with respect to structural parameters of the underlying graph that measure its ``tree-likeness''.

We obtain the following results. We show that \deltaExact\ is W[1]-hard when parameterized by the feedback vertex number of the underlying graph (\cref{thm:W1wrtFVS}). 
To this end, we first give a reduction from \textsc{Multicolored Clique} parameterized by the number of colors~\cite{fellows2009multipleinterval} to a variant of \deltaExact\ where the period $\Delta$ is infinite, that is, when the labeling is non-periodic. We use a special gadget (the ``infinity'' gadget) which allows us to transfer the result to a finite period $\Delta$. The latter construction is independent from the particular reduction we use, and can hence be treated as a reduction from the non-periodic to the periodic setting.

We complement this result by showing that the problem is fixed-parameter tractable (FPT) with respect to the feedback edge number of the underlying graph (\cref{thm:FPTwrtFES}). 
Our FPT algorithm preprocesses the input graph in a suitable way that allows us to efficiently guess desired structures. Then we create integer linear programs (ILPs) (with bounded number of variables) and add constraints to those instances by suitably partitioning the fastest paths among pairs of vertices belonging to a small number of interesting sets. The notion of segment is crucial for adding the necessary and sufficient number of linear constraints. We use Lenstra's Theorem~\cite{Lenstra1983Integer} to solve the ILPs efficiently.

Note that this provides a fairly complete picture of the parameterized computational complexity of \deltaExact\ with respect to structural parameters of the underlying graph that measure ``tree-likeness''. To obtain our results, we prove several properties of fastest temporal paths, which may be of independent interest.

%Due to space constraints, proofs of results marked with $\star$ are (partially) deferred to Appendix.

\subparagraph{Related work.} Graph realization problems on static graphs have been studied since the 1960s. We provide an overview of the literature in the introduction.

To the best of our knowledge, we are the first to consider graph realization problems in the temporal setting. However, many other connectivity-related problems have been studied in the temporal setting~\cite{enright2021deleting,MolterRZ21,klobas2023interference,deligkas2022optimizing,erlebach2021temporal,Flu+19a,Zsc+19,CasteigtsCS22}, most of which are much more complex and computationally harder than their non-temporal counterparts, and some of which do not even have a non-temporal counterpart.

There are two problem settings that share similarities with our setting, which we discuss in more detail in the following.

Several problems have been studied where the goal is to assign labels to (sets of) edges of a given static graph in order to achieve certain connectivity-related properties~\cite{KlobasMMS22,MertziosMS19,akrida2017complexity,enright2021assigning}. The main difference to our problem setting is that in the mentioned works, the input is a graph and the sought labeling is not periodic. Furthermore, the investigated properties are temporal connectivity among all vertices~\cite{KlobasMMS22,MertziosMS19,akrida2017complexity}, temporal connectivity among a subset of vertices~\cite{KlobasMMS22}, or reducing reachability among the vertices~\cite{enright2021assigning}. In all cases the duration of the temporal connections is not of interest.

The only work we are aware of that studies problems on periodic temporal graphs is work on the temporal cops and robber game~\cite{Morawietz021,MorawietzRW20,ErlebachS20}. The question here is (in its basic case) whether a certain number of cops are able to catch a robber, if everybody can move along one edge at every time step and the robber is considered caught if it occupies the same vertex as a cop at the same time.

\subparagraph{Preliminaries and notation.}
We already introduced the most central notion and concepts. There are some additional definitions we need to present our proofs and results which we give in the following.

An interval in $\mathbb N$ from $a$ to $b$ is denoted by $[a,b] = \{ i\in \mathbb N  :  a \leq i \leq b\}$; similarly, $[a] = [1,a]$.
An undirected graph~$G=(V,E)$ consists of a set~$V$ of vertices 
and a set~$E \subseteq V \times V$ of edges.
For a graph~$G$, we also denote by~$V(G)$ and~$E(G)$ the vertex and edge set of~$G$, respectively.
We denote an edge $e \in E$ between vertices $u,v \in V$ as a set $e=\{u,v\}$.
For the sake of simplicity of the representation, an edge $e$ is sometimes also denoted by $uv$. 
A path~$P$ in $G$ is a subgraph of $G$ with vertex set~$V(P)=\{v_1,\dots,v_k\}$ and edge set~$E(P)=\{\{v_i,v_{i+1}\} :  1\leq i<k\}$
(we often represent path~$P$ by the tuple~$(v_1,v_2,\dots,v_k)$).

Let $v_1,v_2,\ldots,v_n$ be the $n$ vertices of the graph $G$. 
For simplicity of the presentation~--and with a slight abuse of notation--~we refer during the paper to the entry $D_{i,j}$ of the matrix $D$ as 
$D_{a,b}$, where $a=v_i$ and $b=v_j$. That is, we put as indices of the matrix $D$ the corresponding vertices of $G$ whenever it is clear from the context.
Many times we also refer to a path $P=(u=v_1, v_2, \dots, v_p=v)$ from $u$ to $v$ in $G$,
as a temporal path in $(G,\lambda)$,
where we actually mean that $(P,\lambda,\Delta)$ is a temporal path with $P$ as its underlying (static) path.
We denote with $d(u,v,P)$ the duration of the fastest temporal path from $u$ to $v$ along path $P$. Note that $d(u,v,P) \geq d(u,v)$.

We remark that a fastest path between two vertices in a temporal graph can be computed in polynomial time~\cite{xuan_computing_2003,Wu2016Efficient}.
%Wu et~al.~\cite{Wu2016Efficient} provide an algorithm that computes the duration of the fastest temporal path from a source vertex $v$, in a temporal graph $(G,\lambda)$, to all other vertices,
%that runs in polynomial-time.
Hence, given a $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$, we can compute in polynomial-time  the matrix $D$
which consists of durations of fastest temporal paths among all pairs of vertices in $(G,\lambda,\Delta)$.

@@

\subparagraph{Organization of the paper.}
In \cref{sec:hardness} we present our hardness results, first the NP-hardness in \cref{sec:nphardness} and then the parameterized hardness in \cref{sec:w1hardness}. In \cref{sec:algos} we present our algorithmic results. First we give in \cref{sec:cyclealgo} a polynomial-time algorithm for the case where the underlying graph is a tree. In \cref{sec:FPT} we generalize this and present our FPT result, which is the main result in the paper. In \cref{sec:cyclealgo} we provide a simple polynomial-time algorithm for the case where the underlying graph is a cycle. Finally, we conclude in \cref{sec:conclusion} and discuss some future work directions.

\section{Hardness results for \deltaExact}\label{sec:hardness}
In this section we present our main computational hardness results. In \cref{sec:nphardness} we show that \deltaExact\ is NP-hard even for constant $\Delta$. In \cref{sec:w1hardness} we investigate the parameterized computational hardness of \deltaExact\ with respect to structural parameters of the underlying graph. We show that \deltaExact\ is W[1]-hard when parameterized by the feedback vertex number of the underlying graph.

\subsection{NP-hardness of \deltaExact}\label{sec:nphardness}
In this section we prove that in general it is NP-hard to determine a $\Delta$-periodic temporal graph $(G,\lambda)$ respecting a duration matrix $D$,
even if $\Delta$ is a small constant.

\begin{theorem}\label{thm:NPhardness}
	\deltaExactLong\ is \NP-hard for all $\Delta \geq 3$.
\end{theorem}

\begin{proof}
	We present a polynomial-time reduction from the NP-hard problem NAE 3-SAT~\cite{Schaefer1978complexity}. Here we are given a formula $\phi$ that is a conjunction of so-called NAE (not-all-equal) clauses, where each clause contains exactly 3 literals (with three distinct variables).
    A NAE clause evaluates to \textsc{true} if and only if not all of its literals are equal, that is, at least one literal evaluates to \textsc{true} and at least one literal evaluates to \textsc{false}.
    We are asked whether $\phi$ admits a satisfying assignment.

    Given an instance $\phi$ of NAE 3-SAT, we construct an instance $(D,\Delta)$ of \deltaExact\ as follows.

 We start by describing the vertex set of the underlying graph $G$ of $D$.
\begin{itemize}
\item For each variable $x_i$ in $\phi$, we create three variable vertices $x_i, x_i^T, x_i^F$.
\item For each clause $c$ in $\phi$, we create one clause vertex $c$.
\item We add one additional super vertex $v$.
\end{itemize}
Next, we describe the edge set of $G$.
\begin{itemize}
\item For each variable $x_i$ in $\phi$ we add the following five edges: 
$x_i x_i^T, x_i x_i^F, x_i^T x_i^F, x_i^T v$ and $ x_i^Fv$.
\item For each pair of variables $x_i,x_j$ in $\phi$ with $i \neq j$ we add the following four edges: 
$x_i^T x_j^T, x_i^Tx_j^F, x_i^Fx_j^T$ and $x_i^Fx_j^F$.
\item For each clause $c$ in $\phi$ we add one edge for each literal. Let $x_i$ appear in $c$. If $x_i$ appears non-negated in $c$ we add edge $c x_i^T$. 
If $x_i$ appears negated in $c$ we add edge $c x_i^F$.
\end{itemize}
This finishes the construction of $G$.
For an illustration see~\cref{fig:NP-example}.

%
\begin{figure}[t]
	\noindent
	\makebox[\textwidth]{
		\centering
\begin{tikzpicture}%[xscale=2]

	%%Variable X1
	\node (x1t) at (1,0) {};
	\node (x1f) at (3,0) {};

	%%Variable X2
	\node (x2t) at (5,0) {};
	\node (x2f) at (7,0) {};

	%%Variable x3
	\node (x3t) at (9,0) {};
	\node (x3f) at (11,0) {};

	%%Variable x4
	\node (x4t) at (13,0) {};
	\node (x4f) at (15,0) {};

	%% edges among xt, xf vertices
	%x1 to all
	\draw [dotted,gray] (x1t) [out=30,in=150] to (x2t);
	\draw [dotted,gray] (x1t) [out=30,in=150] to (x3t);
	\draw [dotted,gray] (x1t) [out=30,in=150] to (x4t);
	\draw [dotted,gray] (x1t) [out=30,in=150] to (x2f);
	\draw [dotted,gray] (x1t) [out=30,in=150] to (x3f);
	\draw [dotted,gray] (x1t) [out=30,in=150] to (x4f);

	\draw [dotted,gray] (x1f) [out=30,in=150] to (x2t);
	\draw [dotted,gray] (x1f) [out=30,in=150] to (x3t);
	\draw [dotted,gray] (x1f) [out=30,in=150] to (x4t);
	\draw [dotted,gray] (x1f) [out=30,in=150] to (x2f);
	\draw [dotted,gray] (x1f) [out=30,in=150] to (x3f);
	\draw [dotted,gray] (x1f) [out=30,in=150] to (x4f);
	%x2 to all
	\draw [dotted,gray] (x2t) [out=30,in=150] to (x3t);
	\draw [dotted,gray] (x2t) [out=30,in=150] to (x4t);
	\draw [dotted,gray] (x2t) [out=30,in=150] to (x3f);
	\draw [dotted,gray] (x2t) [out=30,in=150] to (x4f);

	\draw [dotted,gray] (x2f) [out=30,in=150] to (x3t);
	\draw [dotted,gray] (x2f) [out=30,in=150] to (x4t);
	\draw [dotted,gray] (x2f) [out=30,in=150] to (x3f);
	\draw [dotted,gray] (x2f) [out=30,in=150] to (x4f);
	%x3 to all
	\draw [dotted,gray] (x3t) [out=30,in=150] to (x4t);
	\draw [dotted,gray] (x3t) [out=30,in=150] to (x4f);

	\draw [dotted,gray] (x3f) [out=30,in=150] to (x4t);
	\draw [dotted,gray] (x3f) [out=30,in=150] to (x4f);

	%%Variable X1
	\node[vert,label=left:$x_1^T$] (x1t) at (1,0) {};
	\node[vert,label={[fill=white]right:$x_1^F$}] (x1f) at (3,0) {};
	\node[vert,label=above:$x_1$] (x1) at (2,1) {};
	\draw (x1t) -- (x1f) -- (x1) -- (x1t);

	%%Variable X2
	\node[vert,label={[fill=white]left:$x_2^T$}] (x2t) at (5,0) {};
	\node[vert,label={[fill=white]right:$x_2^F$}] (x2f) at (7,0) {};
	\node[vert,label={[fill=white]above:$x_2$}] (x2) at (6,1) {};
	\draw (x2t) -- (x2f) -- (x2) -- (x2t);

	%%Variable x3
	\node[vert,label={[fill=white]left:$x_3^T$}] (x3t) at (9,0) {};
	\node[vert,label={[fill=white]right:$x_3^F$}] (x3f) at (11,0) {};
	\node[vert,label={[fill=white]above:$x_3$}] (x3) at (10,1) {};
	\draw (x3t) -- (x3f) -- (x3) -- (x3t);
